---
title: "Testing the GrowthData dataset in the hdm package"
output: webexercises::webexercises_default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(webexercises)
```


```{r, echo = FALSE, results='asis'}
# Uncomment to change widget colours:
#style_widgets(incorrect = "goldenrod", correct = "purple")
```



# Introduction

```{r}
# load packages
library(hdm)
library(tidyverse)
```

I have been trying to provide my students with some instructions for using your package. I have been struggling to work out some issues about the GrowthData dataset you helpfully provide through the package. In the help function (`?GrowthData`) you have two links.

[NBER link](https://www2.nber.org/pub/barro.lee/) and [Bristol link](https://www.bristol.ac.uk//Depts//Economics//Growth//barlee.htm)

The former does not seem available any more, but the 2nd from Bristol is still available. I was trying to reconstruct which the 90 countries in the GrowthData dataset are. I reconstructed the data from the Bristol source. As I was doing that I came across two questions I could not answer

1.	In the help for the dataset you say that there were 90 countries for which you had complete data for the 62 covariates. When I try to see how many countries do have complete data from the Bristol dataset 

2. I cannot reconstruct the `Outcome` variable in `GrowthData`.

# Demo

This is to demonstrate my question and hopefully make it easier for you to understand what my question is and if there as an issue to perhaps diagnose the issue.

Let's load the GrowthData data provided by the `hdm` package.

```{r}
data(GrowthData)
str(GrowthData)
```

## Reconstruction of the full Barro and Lee dataset 

I used the data on the Bristol link to reconstruct the full growth dataset. The resulting dataset is available in [GrowthData_complete.csv](https://raw.githubusercontent.com/datasquad/ECLR/gh-pages/data/BARLEE/GrowthData_complete.csv). The construction of the dataset is documented here (I wrote this as an exercise for my students) from [here](https://datasquad.github.io/ECLR/Ex/DataManagementExample_GrowthData.html).

```{r}
GD_complete <- read.csv("https://raw.githubusercontent.com/datasquad/ECLR/gh-pages/data/BARLEE/GrowthData_complete.csv")
```

That data frame has data for 138 countries and 558 variables. As outlined in the help call for `GrowthData` it is a subset of the complete dataset, in particular it says "The number of covariates is p=62. The number of complete observations is 90."

It could of course be that the defunct NBER link contained the exact `GrowthData` dataset, but as that link is defunct I cannot reconstruct this.

## Question 1, which countries?

Object `GrowthData` does not contain a country identifier. That is, of course, totally understandable as it only serves as an illustration for the package functionality, but I was curious and wanted to give my students the task to match the countries.

My first attempt was to check whether `GD_complete` has 90 observations with complete observations for the 62 variables in `GrowthData`.

```{r}
sel_vars <-names(GrowthData)[3:63] # excluding Outcome and intercept
GD_test1 <- GD_complete %>% select(COUNTRY.NAME, sel_vars)
GD_test1 <- GD_test1 %>% drop_na()
dim(GD_test1)
```

This only produces 41 countries.

Then I attempted to match on the basis of the outcome variable, but realised that I think that the outcome variable is not well defined, more on that in Question 2. I therefore decided to match countries on population data, `pop65` (the population of the country in 1965 in thousands).

```{r}
GrowthData_match_pop <- GrowthData %>% select(pop65)
GrowthData_match_pop$GDhdm <- "pop65 match in GrowthData"
GD_test2 <- GD_complete %>% select(COUNTRY.NAME, sel_vars)
GD_test2 <- merge(GD_test2,GrowthData_match_pop)
dim(GD_test2)
```

This again selects 41 countries where the population info coincides between the two data files. Are they the same countries?

```{r}
GD_test1$COUNTRY.NAME[!(GD_test1$COUNTRY.NAME %in% GD_test2$COUNTRY.NAME)]
GD_test2$COUNTRY.NAME[!(GD_test2$COUNTRY.NAME %in% GD_test1$COUNTRY.NAME)]
```

They are the identical set of countries. So, how to identify the remaining 49 countries. Let's look at one of these countries, the one with the biggest population.

```{r}
GrowthData_notmatched <- GrowthData %>%  filter(!(pop65 %in% GD_test1$pop65)) %>% 
  filter((pop65 == min(pop65)) | (pop65 == max(pop65)))
GrowthData_notmatched$pop65

```

So there is a non-matched country with a population in 1965 of 1,968,000 and 620,701,000. Really the large one should be China (and China is not one of the matched countries) Let us check the population information from `GD_complete` for unmatched countries

```{r}
GD_notmatched <- GD_complete %>%  filter(!(pop65 %in% GD_test1$pop65)) %>% 
  filter(!is.na(pop65)) %>% 
  select(COUNTRY.NAME, pop65) %>% arrange(-pop65)

head(GD_notmatched,3)
```

The largest is indeed China but with a population difference of about 100m to the population in in `GrowthData` (gapminder gives a population of 723m for China in 1965). 

```{r}
tail(GD_notmatched)
```

Clearly there are much smaller countries than the smallest unmatched one in `GrowthData`

```{r}
GD_notmatched[52:67,]
```
Jordan is the country with the population closest to 1968, the smallest unmatched population in `GrowtData`. 

We can see whether other characteristics match.

```{r}
GD_Jordan <- GD_complete %>% select(COUNTRY.NAME, sel_vars) %>% 
  filter(pop65 == 1962)
GrowthData_Jordan <- GrowthData %>% select(sel_vars) %>% 
  filter(pop65 == 1968)

GD_Jordan$s65
GrowthData_Jordan$s65

GD_Jordan$lifee065
GrowthData_Jordan$lifee065
```

Even if `GrowthData` reports life expectancy in logs, this is not the same as log(50.3) = 3.918. 

So, it appears as if `GrowthData` draws from some data that are not contained in `GD_complete`. I assume that the dataset that was available from the NBER link may have contained some updated data.

## Question 2. Outcome variable

I tried to reconstruct what the outcome variable is. In the help function it says that it is a growth rate calculated as a log difference. But the help function is somewhat vague about the period and whether it is an annualised growth rate.

To reconstruct this I looked at the data for the UK, which is one of the matched countries identified from the 1965 population of 54,350,000. 

```{r}
GrowthData_UK <- GrowthData %>%  filter(pop65 == 54350)
GrowthData_UK$Outcome
GrowthData_UK$gdpsh465
```

The per capita GDP is in log form, it translates to 5609.

```{r}
sel_vars_gdp1 <- names(GD_complete)[grepl("gdpsh4",names(GD_complete))]
sel_vars_gdp2 <- names(GD_complete)[grepl("grsh4",names(GD_complete))]

GD_UK <- GD_complete %>% filter(COUNTRY.NAME == "United Kingdom") %>% 
  select(sel_vars_gdp1,sel_vars_gdp2)
GD_UK
```

So you can see that the 65 GDP is indeed 5609. The annualised growth rate from 65 to 70 is (`grsh42`) $log(6319/5609)/5 = 0.023828$. Annualised the growth rate between 65 and 70 was 2.38%. This then begs the question what the outcome variable in `GrowthData` for the UK is. It is given with a value of 0.04718. This is around twice as large as the annualised growth rates calculated above.

The help function says "growth rates in GDP per capita for the periods 1965-1975 and 1975-1985". This would suggest the growth rate between 65 and 85. This is

```{r}
log(8665/5609)
```

or in annualised form

```{r}
log(8665/5609)/20
```

The complete datasets contains GDP data from two other sources, but the same issue arises.

So, I am failing to be able to reconstruct the `Outcome` variable.

# Summary

If you make it to the end of this document you have spent some time with this query. I am very grateful that you take this query seriously. If you had any information that would help me understand where I am going wrong I would be very grateful.

Ralf Becker

ralf.becker@manchester.ac.uk